My wonderful paper

### 2024-08-01 _manuscript/_tex/index.tex
add more content

In this paper, a concept called analysis plan is proposed to describe
the logical structure of a data analysis. An analysis plan is a set of
analysis steps plus their expected outcomes. It is a formal
representation of the analysis process and can be used to guide the
analysis process, to communicate and compare the analysis process to
others, and to evaluate the analysis process. The concept of analysis
plan is illustrated with examples and the implications of the concept
for data analysis practice is discussed.

@@

\section{Analysis plan}\label{analysis-plan}

@@

\section{Examples}\label{examples}

@@

\subsection{Linear regression}\label{linear-regression}

Consider a linear regression model to study the effect of PM10 on
mortality (provide context of using PM10 to study mortality). From the
literature, analysts may expect a significant PM10 coefficient in the
linear model from the literature. This is the outcome expectation:
p-value of PM10 coefficient is less than 0.05. There are multiple
factors that can affect the outcome expectation of this analysis, for
example, 1) sample size, 2) model specification, and 3) correlation
structure between variables. Temperature is often an important
confounder to consider in such study (add reference).

To construct the result universe, we can simulate datasets with
different configerations of the above factors. Here, sample sizes of 50,
100, 500, and 1000 are considered. Two model specifications are
included: 1) linear model with PM10 as the only covariate
(\(\text{mortality} \sim \text{PM10}\)), 2) linear model with PM10 and
temperature as covariates
(\(\text{mortality} \sim \text{PM10} + \text{temp}\). A grid-based
approach is used to simulate correlation structure. Reasonable ranges of
correlation between the three variables are

\begin{align*}
\text{cor}(\text{mortality}, \text{PM10}) &\in [-0.01, 0] \\ 
\text{cor}(\text{mortality}, \text{temperature}) &\in [-0.6, -0.2] \\ \text{cor}(\text{PM10}, \text{temperature}) &\in [0.2, 0.6] 
\end{align*}

\includegraphics{index_files/figure-pdf/unnamed-chunk-2-1.pdf}

@@

\captionsetup{labelsep=none}\includegraphics{index_files/figure-pdf/fig-result-universe-1.pdf}

@@

\caption{\label{fig-result-universe}}

@@

Figure~\ref{fig-result-universe} shows that result universe of the
linear regression model with a) colored by whether the p-value of PM10
is significant (less than 0.05), b) highlighting the effect of adding
temperature to the model for a fixed correlation structure, and c)
highlighting the effect of increasing sample size.

>>>>

(background)

In this paper, a concept called analysis plan is proposed to describe
the logical structure of a data analysis. An analysis plan is a set of
analysis steps plus their expected outcomes. It is a formal
representation of the analysis process and can be used to guide the
analysis process, to communicate and compare the analysis process to
others, and to evaluate the analysis process. The concept of analysis
plan is illustrated with examples. The implications of the concept for
data analysis practice is discussed.

@@

The rest of the paper is organized as follows: Section~\ref{sec-plan}
describes the concept of analysis plan in detail.
Section~\ref{sec-examples} provides examples of analysis plan {[}more
details{]}. (need another section here or before examples?)
Section~\ref{sec-conclusion} concludes the paper.

\section{Analysis plan}\label{sec-plan}

@@

\section{Examples}\label{sec-examples}

Three examples are presented to illustrate how the concept of analysis
plan can be applied to data analysis. {[}toy example{]}.
Section~\ref{sec-linear-reg} illustrates how constructing the result
universe in a linear regression model of PM10 on mortality can help
understand the impact of sample size, model specification, and variable
correlation structure on data analysis. {[}example three{]}

@@

\subsection{Linear regression}\label{sec-linear-reg}

Consider a linear regression model to study the effect of PM10 on
mortality (provide context of using PM10 to study mortality). Analysts
may expect a significant (p-value \(\le\) 0.05) PM10 coefficient in the
linear model from the literature. This is the \emph{outcome
expectation}. There are multiple factors that can affect the outcome
expectation of linear regression, which here is called \emph{plan
expectation}, for example, 1) sample size, 2) model specification, and
3) correlation structure between variables. Adequate sample size is
required to achieve the desired power to detect the significance of PM10
on mortality. Temperature is often an important confounder to consider
in such study (add reference). From some domain knowledge, an analyst
may expect that the significance of PM10 coefficient can be attained by
adding temperature to the model. Analysts may also expect certain
correlation structure between PM10, temperature, and mortality, and the
distribution of each variable.

To build the result universe, datasets can be simulated to either meet
and fail these plan expectations, allowing the analysts to observe the
significance of PM10 coefficient. Here, sample sizes of 50, 100, 500,
and 1000 are considered. Two model specifications are included: 1)
linear model with PM10 as the only covariate
(\(\text{mortality} \sim \text{PM10}\)), 2) linear model with PM10 and
temperature as covariates
(\(\text{mortality} \sim \text{PM10} + \text{temp}\)). A grid-based
approach is used to simulate correlation structure. Reasonable ranges of
correlation between the three variables are
\(\text{cor}(\text{mortality}, \text{PM10}) \in [-0.01, 0]\),
\(\text{cor}(\text{mortality}, \text{temperature}) \in [-0.6, -0.2]\),
and \(\text{cor}(\text{PM10}, \text{temperature}) \in [0.2, 0.6]\).

Figure~\ref{fig-result-universe} shows that result universe of the
linear regression model and how a change of decision in one of the plan
expectations above affect the outcome expectation. Panel a) is colored
by the outcome expectation -- whether a significant p-value is found in
the PM10 coefficient. Panel b) shows the effect of adding temperature to
the model and the results show that the significance of PM10 coefficient
can be achieved by adding temperature to the model for a sample size of
500. Panel c) shows that increasing sample size from 50 to 100 enhances
the significance of p-value for PM10 and the significance remains with
further increases in sample size.

A result universe constructed in this example can be presented to
analysts to answer the what-if questions raised in the data analysis.
What if the sample size is increased? What if temperature is added to
the model? What would the results expect to be changed when the
correlation structure is different? For analysts, expectations can be
used as unit tests to divide the result universe, which allows them to
understand the impact of each factor on the outcome expectation.

The result universe also provides a holistic view of how the results
obtained by the analysts are situated in all possible results. This can
be seen as a direct towards trustworthy data analysis, where the
audience of the analysis to exercise their own cognitive model
{[}@grolemund\_cognitive\_2014{]} to evaluate the results reported.

\textsubscript{Source:
\href{https://huizezhang-sherry.github.io/paper-analysis-plan/index.qmd.html}{Article
Notebook}}

@@

\includegraphics{index_files/figure-pdf/fig-result-universe-1.pdf}

@@

\caption{\label{fig-result-universe}The result universe of linear
regression model to study the effect of PM10 on mortality: a) colored by
whether the p-value of PM10is significant (less than 0.05), b) the
effect of adding temperature to the model for a sample size of 500, c)
the effect of increasing sample size for a fixed correlation structure.}

@@

\textsubscript{Source:
\href{https://huizezhang-sherry.github.io/paper-analysis-plan/index.qmd.html}{Article
Notebook}}

\section{Conclusion}\label{sec-conclusion}

\section{References}\label{references}

### 2024-08-01 index.qmd
add more content

---
title: "My wonderful paper"
abstract: "This is the abstract"
bibliography: references.bib
documentclass: jds
format: pdf
preamble: >
  \usepackage{amsfonts,amsmath,amssymb,amsthm}
editor_options: 
  chunk_output_type: console
---

@@

# Analysis plan

@@

# Examples

@@

## Linear regression

Consider a linear regression model to study the effect of PM10 on mortality (provide context of using PM10 to study mortality). Analysts may expect a significant (p-value $\le$ 0.05) PM10 coefficient in the linear model from the literature. There are multiple factors that can affect the outcome expectation of linear regression as such, for example, 1) sample size, 2) model specification, and 3) correlation structure between variables. Temperature is often an important confounder to consider in such study (add reference).

To construct the result universe, we can simulate datasets with different configerations of the above factors. Here, sample sizes of 50, 100, 500, and 1000 are considered. Two model specifications are included: 1) linear model with PM10 as the only covariate ($\text{mortality} \sim \text{PM10}$), 2) linear model with PM10 and temperature as covariates ($\text{mortality} \sim \text{PM10} + \text{temp}$. A grid-based approach is used to simulate correlation structure. Reasonable ranges of correlation between the three variables are

\begin{align*}
\text{cor}(\text{mortality}, \text{PM10}) &\in [-0.01, 0] \\ 
\text{cor}(\text{mortality}, \text{temperature}) &\in [-0.6, -0.2] \\ \text{cor}(\text{PM10}, \text{temperature}) &\in [0.2, 0.6] 
\end{align*}

```{r}
# Generate correlation matrices 
corr_grid <- expand.grid(seq(-0.01, -0.001, by = 0.001), 
                         seq(-0.6, -0.2, 0.05), seq(0.2, 0.6, 0.05))

@@

generate_data <- function(n, mtx, seed = 123) {   
  mu <- c(0, 0, 0)   
  data <- mvrnorm(n, mu, mtx, empirical = TRUE)   
  U <- pnorm(data, mean = 0, sd = 1)   
  set.seed(seed)      
  tibble(mortality = qpois(U[,1], 182), # assume distribution
         pm10 = qnorm(U[,2], mean = 26, sd = 12),           
         temp = qnorm(U[,3], mean = 55, sd = 16))   
  }

@@

```{r}
#| fig.width = 10,
#| fig.height = 5
dt <- res |>   
  mutate(     
    p_value = fit[2,4],     
    coef = fit[2,1],     
    xy_correlation = as.numeric(corr_mtx[1, 2]),     
    xz_correlation = as.numeric(corr_mtx[1, 3]),     
    yz_correlation = as.numeric(corr_mtx[2, 3]),     
    fml = deparse(model) |> as.factor(),     
    expect = ifelse(p_value < 0.05, 1, 0) |> as.factor(),     
    ) |>    
  ungroup()

@@

p0 <- dt |> ggplot() +   
  geom_point(aes(x = p_value , y = coef), color = "grey", size = 0.5) +  
  labs(x = "p-value", y = "coefficient") +   
  scale_color_brewer(palette = "Dark2", direction = -1) +   
  theme_bw() +    
  theme(aspect.ratio = 1, 
        panel.grid.minor = element_blank())

p0 + 
  geom_point(aes(x = p_value , y = coef), size = 0.5) +  
  facet_grid(as.character(model) ~ sample_size) +    
  labs(title = "Coefficient vs. p-value")
```

```{r}
#| label = "fig-result-universe",
#| fig.width = 10,
#| fig.height = 5
df0 <-  dt |> filter(sample_size == 500, xy_correlation == -0.005, 
                     between(xz_correlation, -0.46, -0.43),
                     yz_correlation == 0.4) |> arrange(expect)
#df2 <- df0 |> filter(fml == "mortality ~ pm10 + temp") # test for model spec

p1 <- p0 + 
  geom_point(aes(x = p_value , y = coef, color = expect), size = 0.5) +  
  labs(title = "Coefficient vs. p-value") + 
  theme(legend.position = 'none')

p3 <- p0 + 
  geom_point(aes(x = p_value, y = coef, color = fml), size = 0.5) +
  labs(title = "Effect of adding temperature")

df00 <-  dt |> filter(xy_correlation == -0.005, 
                     between(xz_correlation, -0.46, -0.43),
                     yz_correlation == 0.4) |> arrange(expect)
df4 <- df00 |> filter((fml == "mortality ~ pm10 + temp"))

p5 <- p0 + 
  geom_point(data = df4, aes(x = p_value, y = coef, color = expect), size = 1) +
  geom_path(data = df4, aes(x = p_value, y = coef), color = "black",
            arrow = arrow(type = "open", angle = 30, length = unit(0.1, "inches"))) +
  scale_color_brewer(palette = "Dark2", direction = -1) +
  labs(title = "Effect of sample size")

(p1 | p3 | p5) + plot_annotation(tag_levels = "a") + 
  plot_layout(guides = 'collect') &
  theme(legend.position = 'bottom') 
```

@fig-result-universe shows that result universe of the linear regression model with a) colored by whether the p-value of PM10 is significant (less than 0.05), b) highlighting the effect of adding temperature to the model for a fixed correlation structure, and c) highlighting the effect of increasing sample size.

>>>>

---
title: "My wonderful paper"
abstract: "This is the abstract"
#bibliography: references.bib
documentclass: jds
format: 
  pdf: 
   keep-tex: true
preamble: >
  \usepackage{amsfonts,amsmath,amssymb,amsthm}
editor_options: 
  chunk_output_type: console
---

@@

The rest of the paper is organized as follows: @sec-plan describes the concept of analysis plan in detail. @sec-examples provides examples of analysis plan [more details]. (need another section here or before examples?) @sec-conclusion concludes the paper.

# Analysis plan {#sec-plan}

@@

# Examples {#sec-examples}

Three examples are presented to illustrate how the concept of analysis plan can be applied to data analysis. [toy example]. @sec-linear-reg illustrates how constructing  the result universe in a linear regression model of PM10 on mortality can help understand the impact of sample size, model specification, and variable correlation structure on data analysis. [example three]

@@

## Linear regression {#sec-linear-reg}

Consider a linear regression model to study the effect of PM10 on mortality (provide context of using PM10 to study mortality). Analysts may expect a significant (p-value $\le$ 0.05) PM10 coefficient in the linear model from the literature. This is the *outcome expectation*. There are multiple factors that can affect the outcome expectation of linear regression, which here is called *plan expectation*, for example, 1) sample size, 2) model specification, and 3) correlation structure between variables. Adequate sample size is required to achieve the desired power to detect the significance of PM10 on mortality. Temperature is often an important confounder to consider in such study (add reference). From some domain knowledge, an analyst may expect that the significance of PM10 coefficient can be attained by adding temperature to the model. Analysts may also expect certain correlation structure between PM10, temperature, and mortality, and the distribution of each variable.

To build the result universe, datasets can be simulated to either meet and fail these plan expectations, allowing the analysts to observe the significance of PM10 coefficient. Here, sample sizes of 50, 100, 500, and 1000 are considered. Two model specifications are included: 1) linear model with PM10 as the only covariate ($\text{mortality} \sim \text{PM10}$), 2) linear model with PM10 and temperature as covariates ($\text{mortality} \sim \text{PM10} + \text{temp}$). A grid-based approach is used to simulate correlation structure. Reasonable ranges of correlation between the three variables are $\text{cor}(\text{mortality}, \text{PM10}) \in [-0.01, 0]$, $\text{cor}(\text{mortality}, \text{temperature}) \in [-0.6, -0.2]$, and $\text{cor}(\text{PM10}, \text{temperature}) \in [0.2, 0.6]$.

@fig-result-universe shows that result universe of the linear regression model and how a change of decision in one of the plan expectations above affect the outcome expectation. Panel a) is colored by the outcome expectation -- whether a significant p-value is found in the PM10 coefficient. Panel b) shows the effect of adding temperature to the model and the results show that the significance of PM10 coefficient can be achieved by adding temperature to the model for a sample size of 500. Panel c) shows that increasing sample size from 50 to 100 enhances the significance of p-value for PM10 and the significance remains with further increases in sample size.

A result universe constructed in this example can be presented to analysts to answer the what-if questions raised in the data analysis. What if the sample size is increased? What if temperature is added to the model? What would the results expect to be changed when the correlation structure is different? For analysts, expectations can be used as unit tests to divide the result universe, which allows them to understand the impact of each factor on the outcome expectation.

The result universe also provides a holistic view of how the results obtained by the analysts are situated in all possible results. This can be seen as a direct towards trustworthy data analysis, where the audience of the analysis to exercise their own cognitive model [@grolemund_cognitive_2014] to evaluate the results reported.

```{r sim-linear-res}
# Generate correlation matrices 
corr_grid <- expand.grid(seq(-0.01, -0.001, by = 0.001), 
                         seq(-0.6, -0.2, 0.05), seq(0.2, 0.6, 0.05))

@@

generate_data <- function(n, mtx, seed = 123) {   
  mu <- c(0, 0, 0)   
  data <- mvrnorm(n, mu, mtx, empirical = TRUE)   
  U <- pnorm(data, mean = 0, sd = 1)   
  set.seed(seed)      
  tibble(mortality = qpois(U[,1], 182), # assume distribution
         pm10 = qgamma(U[,2], shape = 6, rate = 0.2),           
         temp = qnorm(U[,3], mean = 55, sd = 16))   
  }

# test for dist fit
# fitdist(pm10$pm10, "gamma") -> pm10_fit
# plot(pm10_fit) # gamma is better than normal for the fit of pm10
#fitdist(pm10$temp, "norm") -> temp_fit
# plot(temp_fit)

@@

```{r}
#| label: fig-result-universe
#| fig-cap: "The result universe of linear regression model to study the effect of PM10 on mortality: a) colored by whether the p-value of PM10is significant (less than 0.05), b) the effect of adding temperature to the model for a sample size of 500, c) the effect of increasing sample size for a fixed correlation structure."
#| fig.width: 12
#| fig.height: 12
dt <- res |>   
  mutate(     
    p_value = fit[2,4],     
    coef = fit[2,1],     
    xy_correlation = as.numeric(corr_mtx[1, 2]),     
    xz_correlation = as.numeric(corr_mtx[1, 3]),     
    yz_correlation = as.numeric(corr_mtx[2, 3]),     
    fml = deparse(model) |> as.factor(),     
    expect = ifelse(p_value < 0.05, 1, 0) |> as.factor(),     
    ) |>    
  ungroup()

@@

p0 <- dt |> ggplot() +   
  geom_point(aes(x = p_value , y = coef), color = "grey", size = 0.5) +  
  labs(x = "p-value", y = "coefficient") +   
  theme_bw() +    
  theme(aspect.ratio = 1, 
        panel.grid.minor = element_blank())

df0 <-  dt |> filter(sample_size == 500, xy_correlation == -0.005, 
                     between(xz_correlation, -0.46, -0.43),
                     yz_correlation == 0.4) |> arrange(expect)
df2 <- df0 |> filter(fml == "mortality ~ pm10 + temp") # test for model spec

p1 <- p0 + 
  geom_point(aes(x = p_value , y = coef, color = expect), size = 0.3) +  
  scale_color_brewer(palette = "Dark2", direction = -1) +
  labs(title = "Coefficient vs. p-value") + 
  theme(legend.position = 'none')

p2 <- p0 + 
  geom_point(data = df0, aes(x = p_value, y = coef, color = expect), size = 2) +
  geom_path(data = df0, aes(x = p_value, y = coef), color = "black",
            arrow = arrow(type = "open", angle = 30, 
                          length = unit(0.1, "inches"))) +
  ggrepel::geom_label_repel(
    data = df0, aes(x = p_value, y = coef, label = fml, color = expect)
    ) +
  scale_color_brewer(palette = "Dark2", direction = -1) +
  labs(title = "Effect of adding temperature")

df00 <-  dt |> filter(xy_correlation == -0.005, 
                     between(xz_correlation, -0.46, -0.43),
                     yz_correlation == 0.4) |> arrange(expect)
df <- df00 |> filter((fml == "mortality ~ pm10 + temp"))

p3 <- p0 + 
  geom_point(data = df, aes(x = p_value, y = coef, color = expect), size = 2) +
  geom_path(data = df, aes(x = p_value, y = coef), color = "black",
            arrow = arrow(type = "open", angle = 30, 
                          length = unit(0.1, "inches"))) +
  ggrepel::geom_label_repel(
    data = df, aes(x = p_value, y = coef, label = sample_size, color = expect)
    ) +
  scale_color_brewer(palette = "Dark2", direction = -1) +
  labs(title = "Effect of sample size")

(p1 | p2) / (p3 | plot_spacer()) + 
  plot_annotation(tag_levels = "a") + 
  plot_layout(guides = 'collect') &
  theme(legend.position = 'bottom') 
```

# Conclusion {#sec-conclusion}

# References
